\bf Model Visualization Techniques for a Social Network Model

### 2017-08-22 paper.Rnw
so close to being done.

We also explore simulations from these different models given parameter values. Using the mean values of from the 1,000 fitted parameter values as the parameters in our models, we simulated 1,000 observations from each of our three models. In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values. From these simulations, we first create a visualization that represents an average network. To do this, we follow the same procedure as in Section~\ref{sec:minds}, counting occurences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times. As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations. In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave. \st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1. Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3. In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2. None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated. In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7. XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.

@@

We also wanted to view the complete microstep process from the first wave, which we condition on in the model, to the second wave. The number of steps taken from wave 1 to wave 2 varies. In the simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103. In the 1,000 simulations, the standard deviation of the number of microsteps was 22.8. In Figure~\ref{fig:allsteps}, we see two simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right. In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way. We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again. There are also some edges that were observed in wave 2 that don't appear at all in the microstep process in a given simulation.

<<alledgests>>=
# get a plot of all microsteps from w1 to w2 by edge
alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
alledges$eid <- 1:nrow(alledges)

@@

ggplot(data = sfmsall4) + 
  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), alpha = alpha), fill = 'navy') +
  scale_alpha_continuous("occurence %") + 
  labs(x = "Microstep No.", y = "Edges (ordered by total # occurences)") + 
  ThemeNoNet + 
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
@

\st{We also combine all 1,000 simulations into a visualization like the one shown in Figure~\ref{fig:allsteps}. We first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations. Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later. We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations. Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep increases. Next, we compute a proportion, which we call the occurence percentage, which is the number of times the edge occurred in a microstep divided by 1,000. Finally, we visualize all this information together in Figure ~\ref{fig:alledgeests}. XX more detail tomorrow XX }

>>>>

We also explore simulations from these different models given parameter values. Using the mean values of from the 1,000 fitted parameter values as the parameters in our models, we simulated 1,000 observations from each of our three models. In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values. From these simulations, we first create a visualization that represents an average network. To do this, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times. As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations. In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave. \st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1. Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3. In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2. None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated. In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7. XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.

@@

We also wanted to view the complete microstep process from the first wave, which we condition on in the model, to the second wave. The number of steps taken from wave 1 to wave 2 varies. In one set of 1,000 simulations from Model 1, the smallest number of steps taken was 58, and the longest was 248, with a mean of 106 and a median of 103. In the 1,000 simulations, the standard deviation of the number of microsteps was 22.8. In Figure~\ref{fig:allsteps}, we see two simulations of the process from wave 1 to wave 2, with wave 1 shown on the left, and wave 2 shown on the right. In each of the three plots, the $y$-axis contains the edges sorted by how often they appear in the networks along the way. We can see that some edges are there in the beginning, but disappear and never come back, while others appear a few steps in, only to dispappear again. There are also some edges that were observed in wave 2 that don't appear at all in the microstep process in a given simulation.

<<alledgests, fig.height = 4, fig.cap="Visualizing all microsteps taken in 1,000 simulations from the model M1. The first wave network is shown at microstep 0, and the second wave of the network is shown as the last microstep for comparison. We see that it is rare for a microstep process to last longer than 150 steps, and also that the edges that appear past the 150th step tend to be in either the first wave or the second wave.">>=
# get a plot of all microsteps from w1 to w2 by edge
alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
alledges$eid <- 1:nrow(alledges)

@@

#add true wave 2 to plot
wave2 <- na.omit(actual2)
wave2$from <- paste0("V", wave2$from)
wave2$to <- paste0("V", wave2$to)
wave2 <- left_join(wave2, alledges)
wave2 <- left_join(wave2, sfmsallorder)

sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1, 
        eid = wave2$eid, 
        count = NA, 
        alpha = 1, 
        from = wave2$from,
        to = wave2$to,
        total = wave2$total, 
        plotorder = wave2$plotorder)

ggplot(data = sfmsall5) + 
  geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alpha)) +
  #scale_alpha_continuous("occurence %") + 
  scale_fill_gradient(low = 'grey90', high = '#0868ac', name = "occurrence %") + 
  labs(x = "Microstep No.", y = "Edges (ordered by total # occurrences)") + 
  ThemeNoNet + 
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank(), panel.grid = element_blank())
@

\st{We also combine all 1,000 simulations into a visualization like the one shown in Figure~\ref{fig:allsteps}. We first assign each possible edge an edge ID number so that we can keep track of it throughout all the microsteps and all the simlations. Then, we count up the total number of times each edge appears in the microstep process in each of the 1,000 simulations for use as an ordering variable later. We also count up the number of times an edge occurs in each microstep number in the 1,000 simulations. Since the number of microsteps in the process varies, the number of times an edge occurs decreases as the microstep number increases. Next, we compute a proportion, which we call the occurrence percentage, which is the number of times the edge occurred in a microstep divided by 1,000. Finally, we visualize all this information together in Figure ~\ref{fig:alledgests}. In this plot, all possible edges are shown, and we see that every one of the $16 \times 15 = 240$ possible edges in the network occurs at some point in the simulation process. We also see, however, that the process struggles to focus in on the edges in the second wave of the data. Ideally, we would like to see more occurrences of the second wave of data generally. About half of the edges in wave two are in the bottom half when ordered by number of occurrences, meaning they aren't appearing as much as they would if the model was truly excellent at capturing the mechanisms of tie change in the network.}
